---
title: "ash depletions check - Yuichi mutation signature"
author: "Kushal K Dey"
date: "9/1/2018"
output: html_document
---

We test the `ash-depletion` model in Logolas on Yuichi's data. The data was collected from
Alexandrov et al paper, where the mutation signatures were computed over around 70,000 somatic
mutations. This is why we take $nP/nQ=10^6$ as a default for this particular set up, and compare 
other choices of nP and nQ and interplay between them. 

```{r warning=FALSE, message=FALSE}
library(Logolas)
```

## Data

```{r}
mat=read.table(paste('../data/Fig4_rawdata/sig_',12,'.txt',sep = ''))
mat1=cbind(t(mat[2:3,1:4]),rep(NA,4),t(mat[4:5,1:4]))
rownames(mat1)=c('A','C','G','T')
colnames(mat1) = c("-2", "-1", "0", "1", "2")
mat2=cbind(rep(NA,6),rep(NA,6),t(mat[1,]),rep(NA,6),rep(NA,6))
colnames(mat2) = c("-2", "-1", "0", "1", "2")
rownames(mat2) = c("C>A", "C>G", "C>T", "T>A", "T>C", "T>G")
table = rbind(mat1, mat2)
table
```

```{r}
cols = RColorBrewer::brewer.pal.info[RColorBrewer::brewer.pal.info$category ==
                                       'qual',]
col_vector = unlist(mapply(RColorBrewer::brewer.pal, cols$maxcolors, rownames(cols)))
col_vector = col_vector[-c(4,5)]
total_chars = c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J", "K", "L", "M", "N", "O",
                "P", "Q", "R", "S", "T", "U", "V", "W", "X", "Y", "Z", "zero", "one", "two",
                "three", "four", "five", "six", "seven", "eight", "nine", "dot", "comma",
                "dash", "colon", "semicolon", "leftarrow", "rightarrow")

set.seed(20)

```

## EDLogo - log

```{r warning=FALSE, message = FALSE}
Logolas::logomaker(table,
                   type = "EDLogo",
                   color_type = "per_symbol",
                   color_seed = 2000,
                   logo_control = list(score = "log",
                                       y_fontsize=10,
                                       control= list(quant=0.5,
                                                     gap_ylab = 2.5,
                                                     round_off=0,
                                                     posbins = 2,
                                                     negbins = 3)))

```

## EDLogo - ash depletion 

### nP = 10^5, nQ = 10^5

```{r warning=FALSE, message=FALSE}
out <- Logolas::logomaker(table,
                   type = "EDLogo",
                   color_type = "per_symbol",
                   color_seed = 2000,
                   logo_control = list(score = "ash",
                                       y_fontsize=10,
                       control= list(quant=0.5,
                                     ash_control = list(nP=10^6, nQ=10^6),
                                     gap_ylab = 2.5,
                                     round_off=0,
                                     posbins = 2,
                                     negbins = 3)))

```


### nP = 10, nQ = 10^6

```{r warning=FALSE, message=FALSE}
out <- Logolas::logomaker(table,
                   type = "EDLogo",
                   color_type = "per_symbol",
                   color_seed = 2000,
                   logo_control = list(score = "ash",
                                       y_fontsize=10,
                       control= list(quant=0.5,
                                     ash_control = list(nP=10, nQ=10^6),
                                     gap_ylab = 2.5,
                                     round_off=0,
                                     posbins = 2,
                                     negbins = 3)))

```

### nP = 10^20, nQ = 10^6

```{r warning=FALSE, message=FALSE}
out <- Logolas::logomaker(table,
                   type = "EDLogo",
                   color_type = "per_symbol",
                   color_seed = 2000,
                   logo_control = list(score = "ash",
                                       y_fontsize=10,
                       control= list(quant=0.5,
                                     ash_control = list(nP=10^20, nQ=10^6),
                                     gap_ylab = 2.5,
                                     round_off=0,
                                     posbins = 2,
                                     negbins = 3)))

```

### nP = 10^6, nQ = 10

```{r warning=FALSE, message=FALSE}
out <- Logolas::logomaker(table,
                   type = "EDLogo",
                   color_type = "per_symbol",
                   color_seed = 2000,
                   logo_control = list(score = "ash",
                                       y_fontsize=10,
                       control= list(quant=0.5,
                                     ash_control = list(nP=10^6, nQ=10),
                                     gap_ylab = 2.5,
                                     round_off=0,
                                     posbins = 2,
                                     negbins = 3)))

```

## EDLogo - preclog

```{r warning=FALSE, message = FALSE}
Logolas::logomaker(table,
                   type = "EDLogo",
                   color_type = "per_symbol",
                   color_seed = 2000,
                   logo_control = list(score = "preclog",
                                       y_fontsize=10,
                                       control= list(quant=0.5,
                                                     gap_ylab = 2.5,
                                                     round_off=0,
                                                     posbins = 2,
                                                     negbins = 3)))

```